Single-fluorophore orientation determination with multiview polarized illumination: modeling and microscope design

### 2017-08-18 paper/paper/paper.tex
Shalin's edits.

\textbf{Spatial/angular variation of emission} techniques image single molecules
and use the distribution of intensity in the back focal plane \cite{lieb2004} or
in the image plane \cite{backer2014} to estimate their orientation. The best
techniques in this category use defocusing or phase masks in the back focal
plane to encode orientation and position information in the image
\cite{agrawal2012}. These techniques can estimate the orientation of molecules
in all orientations and only require a single intensity measurement, but they
require complex reconstruction algorithms, expensive high numerical aperture
(NA) optics, and sensitive detectors.

@@

We can extend these results to light-sheet illumination created by scanning a
laser. If we shine a narrow collimated laser beam onto a galvanometer in the
aperture plane and scan the galvanometer, we create a weakly focused light sheet
in the sample. If we underfill the aperture plane and scan the beam in the
sample with a velocity much less than the beam waist radius divided by the
coherence time, we can ignore the excitation caused by fields aligned with the
optical axis. In this case, molecules in the sample will be excited as if a
single plane wave was incident. Therefore, we can find the absorption efficiency
of a molecule under weakly focused scanned laser illumination by taking the
limit of equation \ref{eq:scalarabs} as $\alpha \rightarrow 0$ (or by plugging
equation \ref{eq:mu} and \ref{eq:excfield} into equation \ref{eq:singleray})
giving
\begin{align}
  \eta_{\text{abs}}^{\text{ray}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to low-NA K\"ohler
illumination with respect to the molecular orientations they excite. Notice that
the absorption efficiency can take its maximum range of values
($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

For each microscope we calculate $\sigma_{\Omega}$ at 10,000
approximately equally spaced points on the unit sphere. A desirable microscope
design will have a small solid-angle uncertainty that is uniform for all
molecule orientations. The most straightforward way to find the location and
scale of the solid-angle uncertainty is to use the mean and variance,
respectively. However, because the solid-angle uncertainty admits extremely
large values that can change the mean and variance dramatically depending on the
sample points, we use the median and median absolute deviation (MAD)---the
median of the data's absolute difference from the median---as robust
alternatives to the mean and variance.

@@

\section{Results}\label{results}
Figure \ref{fig:single-arm} shows our results for single-view designs where a
single objective is used for illumination and detection. We swept through the
illumination and detection NA while keeping the sample exposure constant, and we
found that the lowest median and MAD of the solid-angle uncertainty occurs with
a small illumination NA and a large detection NA where
$\alpha \approx 57.4^{\circ}$. A small illumination NA maximizes the range of
absorption efficiencies, while a large detection NA maximizes the number of
detected photons and the range of the detection efficiencies. Note the relative
importance of illumination and detection NAs---reducing the illumination NA (an
inexpensive modification requiring underfilling the aperture plane) improves the
solid-angle uncertainty much less than increasing the detection NA (an expensive
modification requiring a higher-NA objective). Also note in Figure
\ref{fig:single-arm}b) that single-view designs suffer from high orientation
uncertainty when the molecules are oriented along the optical axis (the
molecules are not efficiently excited or detected), near the transverse plane (it is
ambiguous whether the molecule is oriented above or below the transverse plane),
and near the polarizer orientations (it is ambiguous whether the molecule is on
either side of the polarizer orientation).

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{single-arm}
\caption{Single-view microscope with varying illumination and detection NA. a)
  Schematic of a single-view four-polarization epi-illumination microscope. The
  red arrows indicate the four illumination polarization orientations---one for
  each intensity measurement. b) Solid-angle uncertainty for the microscope in
  a). c) Median of the solid-angle uncertainty as a function of illumination and
  detection NA. d) MAD of the solid-angle uncertainty as a function of
  illumination and detection NA. The microscope in a) and b) is indicated by a
  cross in c) and d).}
\label{fig:single-arm}
\end{figure}

Figures \ref{fig:symmetric-widefield} and \ref{fig:double-arm} show our results
for dual-view symmetric K\"ohler designs. We illuminate from one objective and
detect from the other for four polarization orientations, then we repeat the
polarization orientations with the same objectives in reversed roles. In Figure
\ref{fig:symmetric-widefield} we sweep through the NA of both objectives and the
angle between the objectives while considering steric constraints. Note how
adding a second view removes the high uncertainty region in the transverse
plane. The dual-view microscope still has high uncertainty regions near the
optical axes, but the uncertainty is reduced by almost two orders of magnitude
compared to the single-view microscope. We find that the lowest median of the
solid-angle uncertainty occurs with the largest possible NA objectives. We also
find that increasing NA always lowers the MAD, but orthogonal arms are not
always best. At large NA, it is advantageous to move the objectives together or
against the cover slip. Given the high uncertainty in the transverse plane for a
single-view shown in Figure \ref{fig:single-arm}b), we can see why oblique views
perform better than orthogonal views---oblique views are complementary because
the regions of high orientation uncertainty from each view do not overlap.

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{symmetric-widefield}
\caption{Dual-view symmetric K\"ohler designs with varying NA and angle between
  objectives. a) Schematic of the microscope. We illuminate with the first objective
  (red solid) and detect from the second objective (red dashed). Then we
  illuminate from the second objective (blue solid) and detect from the first
  objective (blue dashed). b) Solid-angle uncertainty for the microscope in
  a). c) Median of the solid-angle uncertainty as a function of NA and the angle
  between the objectives. d) MAD of the solid-angle uncertainty as a function of NA
  and the angle between the objectives. The microscope in a) and b) is indicated
  by a cross in c) and d). e) Median of the solid-angle uncertainty as a
  function of the angle between the objectives when NA=0.6. f) MAD of the
  solid-angle uncertainty as a function of the angle between the objectives when
  NA=0.6. The profile in e) and f) is taken along the orange line in c) and d).}
\label{fig:symmetric-widefield}
\end{figure}

Figure \ref{fig:double-arm} shows our results when we used a dual-view symmetric
orthogonal K\"ohler design and varied the illumination and detection NA. Our
results are similar to the single-view case in Figure \ref{fig:single-arm}. We
find that the best designs use a small illumination NA and a large detection NA
where $\alpha \approx 57.4^{\circ}$. As we discussed in Section
\ref{excitation}, low-NA K\"ohler illumination is approximately equivalent to
light-sheet illumination for the task of estimating the orientation of
molecules. Therefore, the preference for low-NA K\"ohler illumination also
implies that dual-view light-sheet illumination geometries are an excellent
choice for uniformly reconstructing the orientation of single molecules.

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{double-arm}
\caption{Dual-view symmetric orthogonal K\"ohler designs with varying
  illumination and detection NA. a) Schematic of the microscope. b) Solid-angle
  uncertainty for the microscope in a). c) Median of the solid-angle uncertainty
  as a function of illumination and detection NA. d) MAD of the solid-angle
  uncertainty as a function of illumination and detection NA. The microscope in
  a) and b) is indicated by a cross in c) and d).}
\label{fig:double-arm}
\end{figure}

@@

\begin{figure}[htbp]
  \centering\includegraphics[width=\textwidth]{asymmetric-double}
  \caption{Dual-view asymmetric light-sheet illumination designs with varying NA and
    sample exposure asymmetry. a) Schematic of the microscope. b) Solid-angle
    uncertainty for the microscope in a). c) Median of the solid-angle
    uncertainty as a function of NA and sample exposure asymmetry. d) MAD of the
    solid-angle uncertainty as a function of NA and sample exposure
    asymmetry. The microscope in a) and b) is indicated by a cross in c) and
    d).}
\label{fig:asymmetric-double}
\end{figure}

@@

\section{Discussion and Conclusions}\label{discussion}
In this work we have developed a model to predict the intensities measured by
polarized illumination microscopes in a wide range of geometries, developed
metrics to measure the ability of a microscope design to estimate the
orientation of single molecules, and used these metrics to compare a wide
range of microscope designs. Our main result is a short list of design
heuristics that can be used to design polarized-illumination single-molecule
orientation microscopes:
\begin{itemize}
\item All microscopes benefit from a small illumination NA and a large detection
  NA with $\alpha \approx 57.4^{\circ}$.
\item Dual-view microscopes outperform single-view microscopes, and dual-view
  microscopes remove the high uncertainty region in the transverse plane of
  single-view microscopes.
\item High-NA dual-view K\"ohler microscopes are best when the two views are
  oblique. Oblique views are complementary because their high uncertainty
  transverse planes do not overlap.
\item Symmetric NA dual-view orthogonal light-sheet microscopes provide a good
  compromise between cost, solid-angle uncertainty median, and solid-angle
  uncertainty MAD. Asymmetric NA dual-view orthogonal light-sheet microscopes
  can improve the solid-angle uncertainty median or MAD by changing the sample
  exposure ratio between the two views.
\end{itemize}

>>>>

\textbf{Spatial/angular variation of emission} techniques image single molecules
and use the distribution of intensity in the back focal plane \cite{lieb2004} or
in the image plane \cite{backer2014} to estimate their
orientation. Information-theoretically optimal techniques in this category use
defocusing or phase masks in the back focal plane to encode orientation and
position information in the image \cite{agrawal2012}. These techniques can
estimate the orientation of molecules in all orientations and only require a
single intensity measurement, but they require complex reconstruction
algorithms, expensive high numerical aperture (NA) optics, and sensitive
detectors.

@@

We can extend these results to light-sheet illumination created by scanning a
laser. If we shine a narrow, collimated, laser beam onto a galvanometer in the
aperture plane and scan the galvanometer, we create a weakly focused light sheet
in the sample. If we scan the beam in the sample with a velocity much less than
the beam waist radius divided by the coherence time, then we can ignore the
coherence effects of scanning---the light sheet is effectively composed of many
incoherent beams. If we also ensure that the laser beam is weakly focused, we
can ignore the excitation caused by the small fields aligned with the optical
axis. Under slowly-scanned and weakly-focused light-sheet illumination,
molecules in the sample will be excited as if a single plane wave was
incident. Therefore, we can find the absorption efficiency of a molecule in this
case by taking the limit of equation \ref{eq:scalarabs} as
$\alpha \rightarrow 0$ (or by plugging equation \ref{eq:mu} and
\ref{eq:excfield} into equation \ref{eq:singleray}) giving
\begin{align}
  \eta_{\text{abs}}^{\text{ray}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to low-NA K\"ohler
illumination with respect to the molecular orientations they excite. Notice that
the absorption efficiency can take its maximum range of values
($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

For each microscope we calculate $\sigma_{\Omega}$ at 10,000 approximately
equally spaced points on the unit sphere. A desirable microscope design will
have a small solid-angle uncertainty that is uniform for all molecule
orientations. The most straightforward way to find the location and scale of the
solid-angle uncertainty is to use the mean and variance, respectively. However,
the solid-angle uncertainty can take extremely large values, e.g., when a
fluorescent dipole is aligned with the optical axis of a single excitation
objective. Such large values can change the mean and variance dramatically
depending on the sample points, so we use the median and median absolute
deviation (MAD)---the median of the data's absolute difference from the
median---as robust alternatives to the mean and variance.

@@

\section{Results}\label{results}
Figure \ref{fig:single-arm} shows our results for single-view designs where a
single objective is used for illumination and detection. We swept through the
illumination and detection NA while keeping the sample exposure constant, and we
found that the lowest median and MAD of the solid-angle uncertainty occurs with
a small illumination NA and a large detection NA where
$\alpha \approx 65^{\circ}$ (see Figure \ref{fig:single-arm}d) at
NA${}_{\text{det}} \approx 1.2$). A small illumination NA maximizes the range of
absorption efficiencies, while a large detection NA optimizes the number of
detected photons and the range of the detection efficiencies. The optimal
detection NA is slightly larger than $\alpha = 57.4^{\circ}$ because increasing
the detection NA reduces the effect of shot noise. Note the relative importance
of illumination and detection NAs---reducing the illumination NA (an inexpensive
modification requiring underfilling the aperture plane) improves the solid-angle
uncertainty much less than increasing the detection NA (an expensive
modification requiring a higher-NA objective). Also note in Figure
\ref{fig:single-arm}b) that single-view designs suffer from high orientation
uncertainty when the molecules are oriented along the optical axis (the
molecules are not efficiently excited or detected), near the transverse plane
(it is ambiguous whether the molecule is oriented above or below the transverse
plane), and near the polarizer orientations (it is ambiguous whether the
molecule is on either side of the polarizer orientation).

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{single-arm}
\caption{Single-view microscope with varying illumination and detection NA. a)
  Schematic of a single-view four-polarization epi-illumination microscope. The
  red arrows indicate the four illumination polarization orientations---one for
  each intensity measurement. b) Solid-angle uncertainty for the microscope in
  a). c) Median of the solid-angle uncertainty as a function of illumination and
  detection NA. d) MAD of the solid-angle uncertainty as a function of
  illumination and detection NA. The microscope configuration in a) and b) is
  indicated by a cross in c) and d).}
\label{fig:single-arm}
\end{figure}

Figures \ref{fig:symmetric-widefield} and \ref{fig:double-arm} show our results
for dual-view symmetric K\"ohler designs. We illuminate from one objective and
detect from the other for four polarization orientations, then we repeat the
polarization orientations with the same objectives in reversed roles. In Figure
\ref{fig:symmetric-widefield} we sweep through the NA of both objectives and the
angle between the objectives while considering steric constraints. Note how
adding a second view removes the high uncertainty region in the transverse
plane. The dual-view microscope still has high uncertainty regions near the
optical axes, but the uncertainty is reduced by almost two orders of magnitude
compared to the single-view microscope. We find that the lowest median of the
solid-angle uncertainty occurs with the largest possible NA objectives. We also
find that increasing NA always lowers the MAD, but orthogonal arms are not
always optimal. At large NA, it is advantageous to move the objectives together or
against the cover slip. Given the high uncertainty in the transverse plane for a
single-view shown in Figure \ref{fig:single-arm}b), we can see why oblique views
perform better than orthogonal views---oblique views are complementary because
the regions of high orientation uncertainty from each view do not overlap.

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{symmetric-widefield}
\caption{Dual-view symmetric K\"ohler designs with varying NA and angle between
  objectives. a) Schematic of the microscope. We illuminate with the first
  objective (red solid) and detect from the second objective (red dashed). Then
  we illuminate from the second objective (blue solid) and detect from the first
  objective (blue dashed). b) Solid-angle uncertainty for the microscope in
  a). c) Median of the solid-angle uncertainty as a function of NA and the angle
  between the objectives. d) MAD of the solid-angle uncertainty as a function of
  NA and the angle between the objectives. The microscope configuration in a)
  and b) is indicated by a cross in c) and d). e) Median of the solid-angle
  uncertainty as a function of the angle between the objectives when NA=0.6. f)
  MAD of the solid-angle uncertainty as a function of the angle between the
  objectives when NA=0.6. The profile in e) and f) is taken along the purple
  line in c) and d).}
\label{fig:symmetric-widefield}
\end{figure}

Figure \ref{fig:double-arm} shows our results when we used a dual-view symmetric
orthogonal K\"ohler design and varied the illumination and detection NA. Our
results are similar to the single-view case in Figure \ref{fig:single-arm}. We
find that optimal designs use a small illumination NA and a large detection
NA. As we discussed in Section \ref{excitation}, low-NA K\"ohler illumination is
approximately equivalent to light-sheet illumination for the task of estimating
the orientation of molecules. Therefore, the preference for low-NA K\"ohler
illumination also implies that dual-view light-sheet illumination geometries are
an excellent choice for uniformly reconstructing the orientation of single
molecules.

\begin{figure}[htbp]
\centering\includegraphics[width=\textwidth]{double-arm}
\caption{Dual-view symmetric orthogonal K\"ohler designs with varying
  illumination and detection NA. a) Schematic of the microscope. b) Solid-angle
  uncertainty for the microscope in a). c) Median of the solid-angle uncertainty
  as a function of illumination and detection NA. d) MAD of the solid-angle
  uncertainty as a function of illumination and detection NA. The microscope
  configuration in a) and b) is indicated by a cross in c) and d).}
\label{fig:double-arm}
\end{figure}

@@

\begin{figure}[htbp]
  \centering\includegraphics[width=\textwidth]{asymmetric-double}
  \caption{Dual-view asymmetric light-sheet illumination designs with varying NA
    and sample exposure asymmetry. a) Schematic of the microscope. b)
    Solid-angle uncertainty for the microscope in a). c) Median of the
    solid-angle uncertainty as a function of NA and sample exposure
    asymmetry. d) MAD of the solid-angle uncertainty as a function of NA and
    sample exposure asymmetry. The microscope configuration in a) and b) is
    indicated by a cross in c) and d).}
\label{fig:asymmetric-double}
\end{figure}

@@

\section{Discussion and Conclusions}\label{discussion}
In this work we have developed a model to predict the intensities measured by
polarized illumination microscopes in a wide range of geometries, developed
metrics to measure the ability of a microscope design to estimate the
orientation of single molecules, and used these metrics to compare a wide
range of microscope designs. Our main result is a short list of design
heuristics that can be used to design polarized-illumination single-molecule
orientation microscopes:
\begin{itemize}
\item Single-view microscopes benefit from a small illumination NA and a large
  detection NA with $\alpha \approx 65^{\circ}$.
\item Dual-view microscopes outperform single-view microscopes, and dual-view
  microscopes remove the high uncertainty region in the transverse plane of
  single-view microscopes.
\item High-NA dual-view K\"ohler microscopes are optimal when the two views are
  oblique. Oblique views are complementary because their high uncertainty
  transverse planes do not overlap.
\item Symmetric NA dual-view orthogonal light-sheet microscopes provide a good
  compromise between cost, solid-angle uncertainty median, and solid-angle
  uncertainty MAD. Asymmetric NA dual-view orthogonal light-sheet microscopes
  can improve the solid-angle uncertainty median or MAD by changing the sample
  exposure ratio between the two views.
\end{itemize}
